\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2022-02-25 Articles/2020-QuantumLogic/Source/quantumlogic.tex
QuantumLogic: first pass from reviewers

@@

\begin{abstract}
It is generally believed that quantum mechanics entails a rejection of classical logic. The main arguments to support this claim come from the non-commutativity of quantum observables, which allegedly would generate a non-distributive lattice of propositions, and from quantum superpositions, which would entail new rules for quantum disjunctions. In this essay we show that such arguments contain serious flaws.  In particular, it is possible to keep the classical distributivity law in the context of quantum theory once the actual meaning of quantum propositions is unambiguously defined and properly taken into account. We moreover argue that the existence of quantum superpositions does not lead to the definition of a new rule for logical disjunctions. Furthermore, contrary to the received view in quantum logic, we explain that quantum mechanics can generate a distributive lattice of propositions. Finally, the remarkable formal similarities between the logical structures of classical and quantum mechanics are highlighted and discussed. 
\vspace{4mm}

@@

A few years after the publication of this paper, the research in quantum logic saw a notable development in both the scientific and philosophical communities, opening new research programs searching for the correct logical structures to understand and describe the physics of QM (cf.\ \cite{Reichenbach:1944}, \cite{Mackey:1957}, \cite{Finkelstein:1963}, \cite{Kochen:1965}, \cite{Jauch:1969}, \cite{Giuntini:2002}, \cite{DallaChiara:2004}, \cite{Pitowsky:2006}, \cite{Giuntini:2018}, \cite{Svozil:2020}, \cite{Holik:2021}), and intense metaphysical debates regarding the nature of logic itself and its empirical status (cf.\ notably \cite{Quine:1951}, \cite{Putnam:1968}, \cite{Dummett:1976}, \cite{Hallett:1982}, \cite{Weingartner:2004}, \cite{Bacciagaluppi:2009}). In addition, it is worth noting that research in quantum logic is still very prolific to the present day with many interesting applications in the fields of quantum information, computation and cryptography. Therefore, although QL is no longer viewed as a solution to the foundational problems affecting quantum theory\footnote{This conclusion is now accepted among experts; for details see e.g.\ \cite{Bacciagaluppi:2009} and \cite{Giuntini:2002}. Referring to this, the latter authors stated explicitly that ``quantum logics are not to be regarded as a kind of ``clue'', capable of solving the main physical and epistemological difficulties of QT [quantum theory]. This was perhaps an illusion of some pioneering workers in quantum logic'' (\cite{Giuntini:2002}, p.\ 225).}, it remains an important discipline of contemporary physics. In this paper we will neither take a stance in the dispute concerning the ontological status of logic and its laws, nor we will consider recent developments of QL given that they are not strictly relevant to the point we are going to make. Indeed, the present essay is focused on the logical structures common to both quantum and classical mechanics, following the spirit of the former discussion.\footnote{In the present essay we will be concerned only with propositional calculus in the context of non-relativistic QM. First-order and higher-order logics will not be discussed here, nor the logic of relativistic quantum theories.}

@@

On the other hand, we can calculate the truth value of $q_i \vee r_i$ directly and show that it is false. The difficulty here is that before the measurement of $S_y$ only $x$-spin is defined, so how are we going to compute statements about the value of $y$-spin? The solution is simple: while the value of an observable is not always defined, its expectation value always is. The idea is then to translate $q_i$ and $r_i$ into propositions about expectation values, which we already know how to handle, and apply the standard rules. Consider the following statements:
\begin{description}
	\item $q'_i$: ``before the measurement, the expectation value for $y$-spin is $\frac{1}{2} \hbar$'',
	\item $r'_i$: ``before the measurement, the expectation value for $y$-spin is $-\frac{1}{2} \hbar$''.
\end{description}
\noindent Because the spin of the electron along any axis is bounded by $\pm \frac{1}{2} \hbar$, the only way that $q'_i$ is true is if $q_i$ is true. Therefore $q'_i$ and $r'_i$ are respectively equivalent to $q_i$ and $r_i$. Consequently $q_i \vee r_i = q'_i \vee r'_i$. Note that $q'_i \vee r'_i$  is the proposition ``before the measurement, the expectation value for $y$-spin is $\pm \frac{1}{2} \hbar$''. Now, before the measurement of $S_y$ we know that the electron $x$-spin is up. While the $y$-spin value is not fully defined---the state is not an eigenstate of $S_y$---the expectation value is fully defined and it is zero. Zero is different from $\pm \frac{1}{2} \hbar$ and therefore $q_i \vee r_i = \bot$. In fact, since any (non-trivial) superposition of $y$-spin up and down will result in an expectation value strictly between $-\frac{1}{2}\hbar$ and $\frac{1}{2}\hbar$, $q_i \vee r_i$ will be true if and only if either $q_i$ or $r_i$ is true. Therefore it is not the case that the quantum disjunction can be true even if none of the elements is true.

@@

Since $\langle \psi , A \psi \rangle$ expresses the expected value for operator $A$, the proposition shows that the Borel algebra can express statements about the expectation value of observables,\footnote{Technically, position and momentum are not bounded operators. However, one can argue that measured position and momentum are indeed bounded as the acceptance of a detector is always limited. Note that all operators for which the spectral theorem applies can be written as the limit of a sequence of bounded operators.} such as ``the expected position $\langle x \rangle$ of the particle is between $x_0$ and $x_1$.'' This would correspond to the Borel set $\{ \psi \in \mathcal{H} \, | \, \langle \psi , A \psi \rangle \in (x_0, x_1) \}$.\footnote{In most cases, one assumes states to be normalized, which is what we have done here for simplicity. Renormalization would lead to the set $\{ \psi \in \mathcal{H} \, | \, \langle \psi , A \psi \rangle / \langle \psi , \psi \rangle \in (x_0, x_1) \}$, which is still a Borel set since renormalization is a continuous function over its domain $\mathcal{H} \setminus \{ 0 \}$.} Consider, for example, ``the expectation of $z$-spin is zero''. This would correspond to the set $V \subseteq \mathcal{H}$ that includes all states, and only the states, for which the expection of $S_z$ is zero. By Proposition 2, $V$ is a Borel set. In fact, the set $U = {0}$ is a Borel set of $\mathbb{R}$ and therefore $V = F_{S_z}^{-1}(U)$ is a Borel set of $\mathcal{H}$. However, $V$ is not a closed subspace. The set $V$ is not the full set $\mathcal{H}$ since, for example, the state with $z$-spin up is not in the set. However, $x$-spin up and $x$-spin down will be in the set $V$, and their span is the full space $\mathcal{H}$. Therefore $V$ is not a closed subspace. Our proposition, despite being physically interesting, does not correspond to a closed subspace but does correspond to a Borel set.

>>>>

\newcommand\change[1]{{\color{red!75!black} #1}}

@@

\begin{abstract}
\change{At the onset of quantum mechanics, it was speculated that the new theory would entail a rejection of classical logic. The main arguments to support this claim come from the non-commutativity of quantum observables, which allegedly would generate a non-distributive lattice of propositions, and from quantum superpositions, which would entail new rules for quantum disjunctions. While the quantum logic program is not as popular as it once was, the core question remains unsettled: what is the relationship between the logical structures of the two theories? In this essay we show that the original arguments promoting quantum logic contain serious flaws, and that quantum theory does satisfy the classical distributivity law once the full meaning of quantum propositions is properly taken into account. Moreover, we show that quantum mechanics can generate a distributive lattice of proposition, which, unlike the one of quantum logic, includes statements about expectation values which are of undoubtable physical interest. Lastly, we show that the lattice of statistical propositions in classical mechanics follows the same structure, yielding an analogue non-commutative sublattice of classical propositions. This shows that the purported difference between classical and quantum logic stems from a misconstructed parallel between the two theories.}
\vspace{4mm}

@@

A few years after the publication of this paper, the research in quantum logic saw a notable development in both the scientific and philosophical communities, opening new research programs searching for the correct logical structures to understand and describe the physics of QM (cf.\ \cite{Reichenbach:1944}, \cite{Mackey:1957}, \cite{Finkelstein:1963}, \cite{Kochen:1965}, \cite{Jauch:1969}, \cite{Giuntini:2002}, \cite{DallaChiara:2004}, \cite{Pitowsky:2006}, \cite{Giuntini:2018}, \cite{Svozil:2020}, \cite{Holik:2021}), and intense metaphysical debates regarding the nature of logic itself and its empirical status (cf.\ notably \cite{Quine:1951}, \cite{Putnam:1968}, \cite{Dummett:1976}, \cite{Hallett:1982}, \cite{Weingartner:2004}, \cite{Bacciagaluppi:2009}). \change{While the approach is still being actively pursued by a sized community, and despite the many interesting applications in the fields of quantum information, computation and cryptography that are being investigated to the present day, QL is no longer viewed as a solution to the foundational problems affecting quantum theory.\footnote{This conclusion is now accepted among experts; for details see e.g.\ \cite{Bacciagaluppi:2009} and \cite{Giuntini:2002}. Referring to this, the latter authors stated explicitly that ``quantum logics are not to be regarded as a kind of ``clue'', capable of solving the main physical and epistemological difficulties of QT [quantum theory]. This was perhaps an illusion of some pioneering workers in quantum logic'' (\cite{Giuntini:2002}, p.\ 225).} Nevertheless, while it may be reasonable for a physicist to simply discard the approach over practical considerations, such as the lack of a desired result, on philosophical grounds it begs the question: what is the relationship between the logical structures of classical and quantum mechanics? How are they different and how are they similar? What should we think about the core insights that lead to the development of QL? Were they substantially correct, or should they be rejected? Given the fundamental role that logic plays in philosophy, we feel that these questions still must be answered regardless of how one may feel towards QL, or we are bound to make the same mistakes.} In this paper we will neither take a stance in the dispute concerning the ontological status of logic and its laws, nor we will consider recent developments of QL given that they are not strictly relevant to the point we are going to make. Indeed, the present essay is focused on the logical structures common to both quantum and classical mechanics, following the spirit of the former discussion.\footnote{In the present essay we will be concerned only with propositional calculus in the context of non-relativistic QM. First-order and higher-order logics will not be discussed here, nor the logic of relativistic quantum theories.}

@@

On the other hand, we can calculate the truth value of $q_i \vee r_i$ directly and show that it is false. The difficulty here is that before the measurement of $S_y$ only $x$-spin is defined, so how are we going to compute statements about the value of $y$-spin? The solution is simple: while the value of an observable is not always defined, its expectation value always is. The idea is then to translate $q_i$ and $r_i$ into propositions about expectation values, which we already know how to handle, and apply the standard rules. Consider the following statements:
\begin{description}
	\item $q'_i$: ``before the measurement, the expectation value for $y$-spin is $\frac{1}{2} \hbar$'',
	\item $r'_i$: ``before the measurement, the expectation value for $y$-spin is $-\frac{1}{2} \hbar$''.
\end{description}
\noindent \change{If $\rho_i$ is the (potentially) mixed state before the measurement, the expectation value for $y$-spin is given by $\text{Tr}(\rho_i S_y) = \frac{1}{2} \hbar \langle \psi_y^+ | \rho_i | \psi_y^+ \rangle - \frac{1}{2} \hbar \langle \psi_y^- | \rho_i | \psi_y^- \rangle$. Therefore $q'_i$ is true if and only if $\text{Tr}(\rho_i S_y) = \frac{1}{2} \hbar$. Because $\rho_i$ is positive definite, this can only happen if $\langle \psi_y^+ | \rho_i | \psi_y^+ \rangle = 1$ and $\langle \psi_y^- | \rho_i | \psi_y^- \rangle = 0$, which means $\rho_i = | \psi_y^+ \rangle \langle \psi_y^+ |$. This means that the expectation value for $y$-spin is $\frac{1}{2} \hbar$ if and only if the electron was prepared in a pure state of $y$-spin up. In other words, because the spin of the electron along any axis is bounded by $\pm \frac{1}{2} \hbar$, the only way that $q'_i$ is true is if $q_i$ is true. We have that $q'_i$ and $r'_i$ are respectively equivalent to $q_i$ and $r_i$.} Consequently $q_i \vee r_i = q'_i \vee r'_i$. Note that $q'_i \vee r'_i$  is the proposition ``before the measurement, the expectation value for $y$-spin is $\pm \frac{1}{2} \hbar$''. Now, before the measurement of $S_y$ we know that the electron $x$-spin is up. While the $y$-spin value is not fully defined---the state is not an eigenstate of $S_y$---the expectation value is fully defined and it is zero. Zero is different from $\pm \frac{1}{2} \hbar$ and therefore $q_i \vee r_i = \bot$. In fact, since any (non-trivial) superposition of $y$-spin up and down will result in an expectation value strictly between $-\frac{1}{2}\hbar$ and $\frac{1}{2}\hbar$, $q_i \vee r_i$ will be true if and only if either $q_i$ or $r_i$ is true. Therefore it is not the case that the quantum disjunction can be true even if none of the elements is true. \change{The overall point is that the incompatibility of the observables does indeed introduce a problem because it is not clear how to evaluate propositions that mix different variables. The use of equivalent propositions over expectation values helps us solve the ambiguities.}

@@

Since $\langle \psi , A \psi \rangle$ expresses the expected value for operator $A$, the proposition shows that the Borel algebra can express statements about the expectation value of observables,\footnote{Technically, position and momentum are not bounded operators. However, one can argue that measured position and momentum are indeed bounded as the acceptance of a detector is always limited. Note that all operators for which the spectral theorem applies can be written as the limit of a sequence of bounded operators.} such as ``the expected position $\langle x \rangle$ of the particle is between $x_0$ and $x_1$.'' This would correspond to the Borel set $\{ \psi \in \mathcal{H} \, | \, \langle \psi , A \psi \rangle \in (x_0, x_1) \}$.\footnote{In most cases, one assumes states to be normalized, which is what we have done here for simplicity. Renormalization would lead to the set $\{ \psi \in \mathcal{H} \, | \, \langle \psi , A \psi \rangle / \langle \psi , \psi \rangle \in (x_0, x_1) \}$, which is still a Borel set since renormalization is a continuous function over its domain $\mathcal{H} \setminus \{ 0 \}$.} Consider, for example, ``the expectation of $z$-spin is zero''. This would correspond to the set $V \subseteq \mathcal{H}$ that includes all states, and only the states, for which the expection of $S_z$ is zero. By Proposition 2, $V$ is a Borel set. In fact, the set $U = {0}$ is a Borel set of $\mathbb{R}$ and therefore $V = F_{S_z}^{-1}(U)$ is a Borel set of $\mathcal{H}$. However, $V$ is not a closed subspace. The set $V$ is not the full set $\mathcal{H}$ since, for example, the state with $z$-spin up is not in the set. However, $x$-spin up and $x$-spin down will be in the set $V$, and their span is the full space $\mathcal{H}$. Therefore $V$ is not a closed subspace. Our proposition, despite being physically interesting, does not correspond to a closed subspace but does correspond to a Borel set.\footnote{\change{Note that we are not claiming that all Borel sets can be constructed in this way or that all Borel sets correspond to physically interesting propositions. First of all, it is not true: $\{ \psi \}$ and $\{ e^{\imath \theta} \psi\}$ are different Borel sets, yet they are not physically distinguishable since a difference in absolute phase is not physically relevant. This can be fixed by considering the Borel sets of the projective space. Second, to really examine the physicality of all Borel sets we need a ``general theory of experimental logic'', of the type provided by [cite kevin kelly] or [cite carcassi aidala], which would go well beyond the present discussion. With appropriate caveats (e.g. the Hilbert space must be separable), one can say that the Borel sets corresponds to propositions that can be associated to an experimental test (what [carcassi aidala] call theoretical statements) which may or may not terminate in any or all cases. We leave this discussion for another work.}}
